---
title: "Diabetes Risk Factors in US Adults"
subtitle: "An Exploratory Analysis using NHANES Open-Source Data"
author: "fatemeh.torabi@pace.cam.ac.uk"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.align="center")

# Chart palette: dark blue, red, grey
COL_NAVY  <- "#1a3a5c"
COL_RED   <- "#c0392b"
COL_GREY  <- "#7f8c8d"
COL_LGREY <- "#bdc3c7"

pkgs <- c("NHANES","tidyverse","gtsummary","plotly","broom","scales","forcats","DT")
invisible(lapply(pkgs, function(p) {
  if (!requireNamespace(p, quietly=TRUE))
    install.packages(p, repos="https://cloud.r-project.org")
}))
library(NHANES); library(tidyverse); library(gtsummary)
library(plotly);  library(broom);    library(scales)
library(forcats); library(DT)
```

---

## 1. Background

### 1.1 The global burden of diabetes

Type 2 diabetes mellitus is one of the most prevalent chronic non-communicable diseases
worldwide. According to the International Diabetes Federation (IDF), approximately
**537 million adults** were living with diabetes in 2021, projected to rise to 783 million
by 2045. In the United States alone, the CDC estimates that **37.3 million Americans**
(~11.3%) have diabetes, with 96 million more classified as pre-diabetic.

People living with diabetes face elevated risks of cardiovascular disease, stroke, chronic
kidney disease, retinopathy, neuropathy, and lower-limb amputations. The American Diabetes
Association estimated the total cost of diagnosed diabetes in the US at **$327 billion**
in 2017.

### 1.2 Risk factors: what do we know?

- **Age** — insulin resistance and beta-cell dysfunction increase progressively with age.
- **BMI** — excess adiposity is one of the strongest modifiable risk factors; obese
  individuals face a 7–10× higher risk than those at healthy weight.
- **Physical inactivity** — sedentary behaviour impairs glucose uptake and insulin
  sensitivity; regular activity reduces incident diabetes risk.
- **Smoking** — current smokers have ~30–40% greater risk than non-smokers, mediated
  through systemic inflammation and oxidative stress.
- **Socioeconomic status** — lower income limits access to nutritious food, recreational
  facilities, and preventive healthcare.
- **Race/ethnicity** — well-documented disparities exist across population groups,
  reflecting both genetic susceptibility and structural inequalities.

### 1.3 Research question

> **Research Question:** *Which sociodemographic and lifestyle factors are independently
> associated with self-reported diabetes diagnosis in US adults, after adjusting for
> potential confounders?*

### 1.4 Data source: NHANES

The **National Health and Nutrition Examination Survey (NHANES)** combines interviews and
standardised physical examinations in a nationally representative probability sample.
The `NHANES` R package provides cleaned data from the 2009–2012 survey cycles (10,000
observations, 76 variables), freely available on CRAN with no registration required.

---

## 2. Data Preparation

### 2.1 Raw data preview

The table below shows a 200-row random sample of raw NHANES restricted to analysis
variables. Notice that `PhysActive`, `SmokeNow`, and `HHIncomeMid` contain `NA` values
arising from NHANES survey routing — the cleaning steps below quantify this.

```{r raw-preview}
data(NHANES)
NHANES %>%
  select(ID, Gender, Age, Race1, BMI, PhysActive, SmokeNow, HHIncomeMid, Diabetes) %>%
  sample_n(200) %>%
  datatable(
    caption="Table 0. Raw NHANES data (200-row sample, analysis variables only)",
    rownames=FALSE, filter="top", extensions="Buttons",
    options=list(pageLength=10, dom="Bfrtip",
                 buttons=c("copy","csv","excel"), scrollX=TRUE)
  )
```

### 2.2 Cohort construction & cleaning

| Step | Action |
|------|--------|
| Age restriction | Adults aged 20–79 only |
| De-duplication | One record per participant (`distinct(ID)`) |
| Outcome coding | Diabetes Yes → 1, No → 0; `NA` excluded |
| BMI category | Underweight / Normal / Overweight / Obese |
| Age group | Four bands: 20–39, 40–54, 55–69, 70–79 |
| Income | `HHIncomeMid ÷ 1,000` (USD thousands) |
| Complete-case | Drop rows with `NA` in any analysis variable |

```{r cohort}
cohort_raw <- NHANES %>%
  filter(Age >= 20, Age < 80) %>%
  distinct(ID, .keep_all=TRUE) %>%
  select(ID, Gender, Age, Race1, BMI, PhysActive, SmokeNow, HHIncomeMid, Diabetes)

n_start      <- nrow(cohort_raw)
n_miss_diab  <- sum(is.na(cohort_raw$Diabetes))
n_miss_bmi   <- sum(is.na(cohort_raw$BMI))
n_miss_phys  <- sum(is.na(cohort_raw$PhysActive))
n_miss_smoke <- sum(is.na(cohort_raw$SmokeNow))
n_miss_inc   <- sum(is.na(cohort_raw$HHIncomeMid))

cohort <- cohort_raw %>%
  mutate(
    Diabetes   = if_else(Diabetes == "Yes", 1L, 0L),
    PhysActive = factor(PhysActive, levels=c("Yes","No"),
                        labels=c("Active","Inactive")),
    SmokeNow   = factor(SmokeNow, levels=c("Yes","No"),
                        labels=c("Current smoker","Non-smoker")),
    Gender     = factor(Gender),
    Race1      = factor(Race1),
    AgeGroup   = cut(Age, breaks=c(20,40,55,70,80),
                     labels=c("20–39","40–54","55–69","70–79"), right=FALSE),
    BMI_cat    = cut(BMI, breaks=c(0,18.5,25,30,Inf),
                     labels=c("Underweight","Normal","Overweight","Obese"),
                     right=FALSE),
    Income_k   = HHIncomeMid / 1000
  ) %>%
  filter(!is.na(Diabetes), !is.na(BMI), !is.na(PhysActive),
         !is.na(SmokeNow), !is.na(HHIncomeMid), !is.na(Race1))

n_final    <- nrow(cohort)
n_excluded <- n_start - n_final

tibble::tribble(
  ~Stage,                              ~N,
  "Adults 20–79 after age filter",     n_start,
  "Missing: Diabetes outcome",         n_miss_diab,
  "Missing: BMI",                      n_miss_bmi,
  "Missing: Physical activity",        n_miss_phys,
  "Missing: Smoking status",           n_miss_smoke,
  "Missing: Household income",         n_miss_inc,
  "────────────────────────────────", NA_integer_,
  "Final analysis-ready cohort",       n_final
) %>%
  knitr::kable(caption="Cleaning flow — records excluded at each stage",
               col.names=c("Stage","N excluded / remaining"))
```

The largest source of missingness is smoking status (`r n_miss_smoke` records), a
structural NHANES routing feature rather than random non-response. Complete-case analysis
excludes **`r format(n_excluded, big.mark=",")`** participants, leaving a final cohort of
**`r format(n_final, big.mark=",")`**.

### 2.3 Analysis-ready cohort preview

```{r cohort-preview}
cohort %>%
  select(ID, Gender, AgeGroup, Race1, BMI, BMI_cat,
         PhysActive, SmokeNow, Income_k, Diabetes) %>%
  head(200) %>%
  datatable(
    caption="Table 0b. Analysis-ready cohort (first 200 rows)",
    rownames=FALSE, filter="top", extensions="Buttons",
    options=list(pageLength=10, dom="Bfrtip",
                 buttons=c("copy","csv","excel"), scrollX=TRUE)
  )
```

> **Final cohort:** `r format(n_final, big.mark=",")` participants ·
> `r sum(cohort$Diabetes)` diabetes cases
> (`r sprintf("%.1f%%", 100*mean(cohort$Diabetes))` crude prevalence) ·
> `r format(n_excluded, big.mark=",")` excluded due to missing data.

---

## 3. Cohort Summary (TableOne)

```{r tableone}
cohort %>%
  select(Gender, AgeGroup, BMI_cat, PhysActive, SmokeNow, Income_k, Diabetes) %>%
  mutate(Diabetes = factor(Diabetes, labels=c("No Diabetes","Diabetes"))) %>%
  tbl_summary(
    by        = Diabetes,
    label     = list(Gender~"Sex", AgeGroup~"Age group", BMI_cat~"BMI category",
                     PhysActive~"Physical activity", SmokeNow~"Current smoker",
                     Income_k~"Household income (USD thousands)"),
    statistic = list(all_continuous()~"{median} ({p25}, {p75})",
                     all_categorical()~"{n} ({p}%)"),
    missing   = "no"
  ) %>%
  add_p() %>% add_overall() %>% bold_labels() %>%
  modify_caption("**Table 1. Cohort characteristics by diabetes status**")
```

**Interpretation:** Every variable differs significantly between groups (all p < 0.05),
confirming that diabetes is not randomly distributed. Key patterns:

- **Age** — participants with diabetes are substantially older; the 55–69 and 70–79 bands
  are heavily over-represented in the diabetes group, consistent with cumulative metabolic
  burden over time.
- **BMI** — the obese category accounts for a markedly higher share of diabetes cases,
  reflecting the adiposity–insulin resistance pathway.
- **Physical activity** — a greater proportion of inactive participants have diabetes,
  suggesting sedentary behaviour is associated with higher disease burden.
- **Household income** — median income is lower in the diabetes group, consistent with
  socioeconomic gradients in chronic disease.
- **Smoking** — distributional differences are present though direction is nuanced, partly
  due to survivorship effects (heavier smokers may have died before the survey).

These univariate differences motivate the multivariable model in Section 5, which
disentangles the independent contribution of each factor.

---

## 4. Exploratory Visualisations

### 4.1 Age distribution by diabetes status

```{r plot-age, fig.height=4}
cohort %>%
  mutate(Status = if_else(Diabetes==1,"Diabetes","No Diabetes")) %>%
  plot_ly(x=~Age, color=~Status,
          colors=c("No Diabetes"=COL_NAVY, "Diabetes"=COL_RED),
          type="histogram", alpha=0.70, nbinsx=30) %>%
  layout(barmode="overlay",
         title=list(text="Age Distribution by Diabetes Status",
                    font=list(size=15)),
         xaxis=list(title="Age (years)"),
         yaxis=list(title="Count"),
         legend=list(title=list(text="Status")))
```

**Interpretation:** The no-diabetes group (dark blue) is broadly distributed across adult
age bands with a roughly bell-shaped peak in mid-life. The diabetes group (red) shows a
right-skewed distribution concentrated in the 50–70 age range, with very few cases under
40. This rightward shift highlights age as one of the dominant drivers of diabetes
prevalence. The near-absence of diabetes below age 35 is consistent with the typically
long prodromal period of type 2 diabetes, during which insulin resistance develops
silently before clinical diagnosis.

---

### 4.2 BMI by physical activity and diabetes

```{r plot-bmi, fig.height=4}
cohort %>%
  mutate(Status = if_else(Diabetes==1,"Diabetes","No Diabetes")) %>%
  plot_ly(x=~PhysActive, y=~BMI, color=~Status,
          colors=c("No Diabetes"=COL_NAVY, "Diabetes"=COL_RED),
          type="box", boxpoints="outliers") %>%
  layout(boxmode="group",
         title=list(text="BMI by Physical Activity and Diabetes Status",
                    font=list(size=15)),
         xaxis=list(title="Physical Activity Status"),
         yaxis=list(title="BMI (kg/m²)"),
         legend=list(title=list(text="Status")))
```

**Interpretation:** Across both activity groups, participants with diabetes (red boxes)
have noticeably higher median BMI than those without (dark blue), underlining the
adiposity–diabetes link. Within the diabetes group, inactive participants have slightly
higher BMI than active ones, suggesting inactivity compounds the metabolic effect of
excess weight. Critically, even among active participants, those with diabetes maintain
higher BMI than their active non-diabetic counterparts — demonstrating that physical
activity alone does not fully offset obesity-related metabolic risk. Exercise and weight
management are complementary, not interchangeable, strategies.

---

### 4.3 Diabetes prevalence by age group and BMI category

```{r plot-prev, fig.height=4}
cohort %>%
  group_by(AgeGroup, BMI_cat) %>%
  summarise(prevalence=100*mean(Diabetes), n=n(), .groups="drop") %>%
  filter(!is.na(BMI_cat)) %>%
  plot_ly(x=~AgeGroup, y=~prevalence, color=~BMI_cat,
          colors=c("Underweight"=COL_LGREY, "Normal"=COL_NAVY,
                   "Overweight"=COL_GREY,   "Obese"=COL_RED),
          type="bar",
          text=~paste0("n = ",n,"<br>Prevalence = ",round(prevalence,1),"%"),
          hovertemplate="%{text}<extra></extra>") %>%
  layout(barmode="group",
         title=list(text="Diabetes Prevalence by Age Group and BMI Category",
                    font=list(size=15)),
         xaxis=list(title="Age Group"),
         yaxis=list(title="Prevalence (%)"),
         legend=list(title=list(text="BMI Category")))
```

**Interpretation:** Two clear gradients are visible simultaneously. Within every BMI
category, prevalence rises consistently with age — the 70–79 band is highest regardless
of weight. Within every age band, obese participants (red bars) carry the highest
prevalence, followed by overweight (grey), normal (dark blue), and underweight (light
grey). The joint effect is starkest in older obese adults, where prevalence exceeds 40%
in some cells. Normal-weight adults maintain considerably lower prevalence across all age
groups, reinforcing healthy weight maintenance as a priority prevention target.

---

## 5. Logistic Regression Model

We fit a multivariable logistic regression with diabetes (1/0) as the binary outcome,
adjusting simultaneously for all covariates. Results are expressed as **odds ratios (OR)**
with 95% confidence intervals. OR > 1 = increased odds; OR < 1 = reduced odds, relative
to the reference category (female sex, White race, physically active, current smoker).

```{r model}
model <- glm(
  Diabetes ~ Age + Gender + BMI + PhysActive + SmokeNow + Income_k + Race1,
  data=cohort, family=binomial(link="logit")
)

model_tbl <- tidy(model, exponentiate=TRUE, conf.int=TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = recode(term,
      "Age"                = "Age (per year)",
      "GenderMale"         = "Male sex",
      "BMI"                = "BMI (per unit)",
      "PhysActiveInactive" = "Physically inactive",
      "SmokeNowNon-smoker" = "Non-smoker (ref: current smoker)",
      "Income_k"           = "Household income (per $1k)",
      "Race1Asian"         = "Race: Asian",
      "Race1Black"         = "Race: Black",
      "Race1Hispanic"      = "Race: Hispanic",
      "Race1Mexican"       = "Race: Mexican",
      "Race1Other"         = "Race: Other / Other White"
    ),
    significant = p.value < 0.05
  )

model_tbl %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(across(c(estimate,conf.low,conf.high), ~round(.x,2)),
         p.value=if_else(p.value<0.001,"<0.001",
                         as.character(round(p.value,3)))) %>%
  rename(Predictor=term, OR=estimate,
         `95% CI Lo`=conf.low, `95% CI Hi`=conf.high, `P-value`=p.value) %>%
  knitr::kable(caption="Table 2. Adjusted odds ratios — multivariable logistic regression")
```

**Interpretation:** After full adjustment, several factors remain independently and
significantly associated with diabetes. Age and BMI show the most consistent effects:
each additional year of age and each one-unit BMI increase both raise the odds of
diabetes independent of all other factors in the model. Physical inactivity raises odds
beyond what BMI alone explains, confirming that exercise confers metabolic benefit
independently of weight. Higher household income is modestly protective. Smoking and male
sex do not reach significance after adjustment, likely reflecting confounding by the
correlated variables already in the model. Racial disparities persist even after
socioeconomic and lifestyle adjustment, pointing to structural and biological factors not
fully captured here.

---

## 6. Forest Plot

```{r forest-plot, fig.height=6.5, fig.width=9}
model_tbl %>%
  mutate(term = fct_reorder(term, estimate)) %>%
  ggplot(aes(x=estimate, y=term, colour=significant)) +
  geom_vline(xintercept=1, linetype="dashed", colour="grey40", linewidth=0.7) +
  geom_errorbarh(aes(xmin=conf.low, xmax=conf.high),
                 height=0.25, linewidth=0.9) +
  geom_point(size=3.5, shape=18) +
  scale_colour_manual(
    values = c("TRUE"=COL_RED, "FALSE"=COL_GREY),
    labels = c("TRUE"="p < 0.05", "FALSE"="p \u2265 0.05"),
    name   = NULL
  ) +
  scale_x_log10(breaks=c(0.5,0.75,1,1.25,1.5,2,3,4),
                labels=label_number(accuracy=0.01)) +
  labs(title    = "Forest Plot: Adjusted Odds Ratios for Diabetes",
       subtitle = "Multivariable logistic regression \u00b7 NHANES adults aged 20\u201379",
       x        = "Odds Ratio (log scale)", y=NULL,
       caption  = "Error bars = 95% CI  \u00b7  Red = p < 0.05  \u00b7  Grey = p \u2265 0.05") +
  theme_minimal(base_size=12) +
  theme(plot.title=element_text(face="bold",size=14),
        plot.subtitle=element_text(colour="grey45",size=11),
        plot.caption=element_text(colour="grey55",size=9),
        panel.grid.major.x=element_line(colour="grey85"),
        panel.grid.minor=element_blank(),
        panel.grid.major.y=element_blank(),
        legend.position="top",
        axis.text.y=element_text(size=10))
```

**Interpretation:** Reading from the bottom (lowest OR) upward:

- **Household income** (red, OR < 1) sits left of the null line — each additional $1,000
  of income is associated with *reduced* odds of diabetes, confirming its protective role.
- **Non-smoker** and **male sex** are grey — their CIs cross OR = 1 after adjustment,
  meaning no independent significant effect is detectable in this sample.
- **BMI** and **physical inactivity** are red and sit right of the null, confirming
  independent increases in odds.
- **Age** has one of the strongest effects on the log scale: even a per-year increment
  yields an OR meaningfully above 1.0, with a very narrow CI — high precision and strong
  statistical significance.
- **Race: Mexican and Hispanic** show the largest ORs among race/ethnicity terms, both
  statistically significant with CIs entirely above 1.0, even after income, BMI, and
  activity are controlled — pointing to residual structural and potential genetic factors.
- **CI width** reflects sample size: wider bars (e.g. Race: Other) signal fewer
  participants in that category and correspondingly greater uncertainty.

---

## 7. Summary & Conclusions

```{r findings, echo=FALSE}
sig     <- model_tbl %>% filter(significant) %>% arrange(estimate)
bmi_or  <- round(filter(model_tbl, grepl("BMI \\(per", term))$estimate, 2)
age_or  <- round(filter(model_tbl, grepl("Age", term))$estimate, 3)
inc_or  <- round(filter(model_tbl, grepl("income", term, ignore.case = TRUE))$estimate, 2)
phys_or <- round(filter(model_tbl, grepl("inactive", term, ignore.case = TRUE))$estimate, 2)
```

This analysis used cross-sectional data from `r format(n_final, big.mark = ",")` US
adults (2009–2012 NHANES) to examine factors associated with diabetes diagnosis.
Multivariable logistic regression identified **`r nrow(sig)` statistically significant
independent predictors**:

| Predictor | Direction | Adjusted OR |
|-----------|-----------|-------------|
| Age (per year) | Risk ↑ | `r age_or` |
| BMI (per unit) | Risk ↑ | `r bmi_or` |
| Physically inactive | Risk ↑ | `r phys_or` |
| Household income (per $1k) | Protective ↓ | `r inc_or` |

**Key messages for public health:**

1. **BMI is the most actionable independent risk factor** — it is modifiable and its
   effect persists after full adjustment. Population-level interventions targeting healthy
   weight through diet quality and built-environment changes are warranted.
2. **Physical activity confers independent benefit beyond weight** — promoting exercise
   should be a priority even in groups where significant weight loss is unlikely.
3. **Socioeconomic disadvantage amplifies risk** — interventions that address food
   insecurity, income inequality, and neighbourhood resources may reduce diabetes
   incidence in lower-income populations.
4. **Racial and ethnic disparities persist after covariate adjustment** — this signals
   the need for targeted screening programmes and further research into structural
   determinants not captured by income alone.

**Limitations:** This is a cross-sectional observational analysis; causality cannot be
inferred from these associations. Complete-case analysis was used, which assumes data are
missing at random — an assumption that may not fully hold for smoking status. Survey
sampling weights were not applied, so point estimates are not precisely nationally
representative. Unmeasured confounders (diet quality, family history, medication use) may
bias associations in unpredictable directions.

---

*Data source: `NHANES` R package (CDC/NCHS, 2009–2012 survey cycles). Analysis restricted
to `r format(n_final, big.mark = ",")` adults aged 20–79 with complete data on all study
variables.*