---
title: "Template report"
author: "fatemeh.torabi@ice.cam.ac.uk"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: cayman
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(knitr)) install.packages("knitr", repos = "http://cran.us.r-project.org")
if(!require(DT)) install.packages("knitr", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(knitr)
library(DT)
```

# Structure of the portfolio

### Abstract 
Provide the reader with a succinct summary of your work

### Introduction
Provide an introduction to you portfolio to reader. 

### Method
covering data access requirements, ethics, metadata and all methodological aspects of your project

### Results
Use this section to showcase the results of your data manipulation that will contribute to the project

### Conclusion
Summaries your findings,discuss them in the context of other similar work or questions and suggestions for future work. Conclude your portfolio with what started your data exploration and what have the data contributed in the decisions for patient care or health service delivery.  


# In text elements
Some examples of having in-text elements as you develop your portfolio are provided here.

* "# Headings"
* "## Subheading 1"
* "### subheading 2"
* "#### subheading 3"

## Hyperlinks

[Healthcare Data Science](https://github.com/CambridgeICE-HDS/MSt-Healthcare-Data-Science)

## Notice box 

::: {.infobox .caution data-latex="{caution}"}

The format to add boxes to your portfolio
:::


## Tables

### In-text table

A template format to add a table into 
the document you can use the following md code structure. 

|Data table    | Coverage     |Area                     |
|--------------|--------------|-------------------------|
|Health survey | 2015         |Self-reported outcomes   |
|EHR           | 2000 onwards |Electronic health records|

# Data tables

## defining a dataframe 

```{r dataframe1, message=FALSE}
CREL <- data.frame(
  Data = c("Health survey","EHR"), 
  Coverage=c(
    "2015","2000 onwards"
  ),
  Area=c(
    "Self-reported outcomes",
    "Electronic Health Records"
  ))

kable(CREL)
```

## Interactive data elements

```{r dataframe2, message=FALSE}
DT::datatable(
  CREL,
  extensions = 'Buttons',
  options = list(
              paging = TRUE,
              searching = TRUE,
              fixedColumns = TRUE,
              autoWidth = TRUE,
              ordering = TRUE,
              dom = 'tB',
              buttons = c('copy', 'excel')
              ),
              class = "display"
)
```
